Assumptions of Physics blueprint: counting evolutions as a foundation for entropy and system/state definition

### 2021-05-14 Overviews/Process entropy/ProcessEntropy.tex
ProcEntropy: first pass done

\subsection{Future: States and state spaces}

In other words, the state space of the system (i.e. the set of all possible complete descriptions of the system and only about that system, independently of the process at hand) is something we need to construct. It is an endpoint, not a starting point. This construction comes with a set of assumptions about the system which are at the foundations of thermodynamics specifically and physics more in general, and they need to be made explicit. Best descriptions of a system at a time, instead, do not require any additional assumptions and therefore are more general concepts.

\textbf{States and independence}. It should be noted that the same description for the same system in different conditions (i.e. in different processes or at different times in the same process) will not in general correspond to the same count of evolutions. Consider a dynamical system with an attractor, such as a damped harmonic oscillator. As time proceeds, more and more evolutions will converge to the attractor. The same finite region around the attractor will correspond to more and more evolutions. In general, a process may introduce correlations between different variables which influences the value of the process entropy for a specific description.

If two systems are independent, however, the evolution of each system does not constrain the evolution of the others. In this case, the count of evolution, and therefore the entropy, of a particular system is only a function of the description of the independent system. The count of evolution factorizes and the entropy becomes additive.

What is crucial to understand here is that \textbf{independence is a requirement for the definition of a system}: we can talk about a system only insofar we have at our disposal a process under which the system is independent for the rest, such that the description of the system is truly a description of the system, with no correlations. We can speak of a chair as its own system because we have circumstances in which we can interact with the chair in a way that does not affect the other chairs or the tables.

Philosophers talk about description of reality that is joint-carving meaning that reality already contains divisions, prior to our investigation. This is an epistemological version of it: nature gives us access to processes for which some systems are independent from the rest. It is the accessibility of these processes that allows us to define systems, without them we cannot divide nature into parts. That is, if an object is really made of parts but the process driving their evolution is such that the correlations can never be eliminated in a controlled fashion, the parts cannot be defined as independent systems.

\section{Current attempt}

The math.

\section{Examples}

\section{Insightful failures}

>>>>

\subsection{System, states and state spaces}

\textbf{Systems}. Lastly we need to answer the following question: how do we go from a set of descriptions to the definition of a system? Under what conditions do the snapshot actually represent states? And why do we do it? The general idea is that, conceptually, we want to study the same object under different processes. That is, we want to say that the same object $X$ can be subjected to different process at different intervals of time, and we are able to switch from one to the other. Note that this is not simply something we would like to do: it is a basic requirement to do science. Process $A$, in fact, can be the reliable preparation of a sample, process $B$ can be some interaction we want to characterize, and process $C$ can be the final measurement. What must happen, then, is that the description of the object at each junction is independent from the choice of the previous/future processes and from other objects. In other words, the description at the junction is about and only about the system at that time. This does not mean that it is the \emph{full} description. It must be, then, that we have at our disposal a process under which the object is independent from the rest. We removed all correlations in time and with other systems. We call an object for which this is possible a \textbf{system}.

\textbf{Independence as requirement}. Therefore \textbf{independence is a requirement for the definition of a system}: we can talk about a system only insofar we have at our disposal a process under which the system is independent for the rest, such that the description of the system is truly a description of the system, with no correlations. We can speak of a chair as its own system because we have circumstances in which we can interact with the chair in a way that does not affect the other chairs or the tables. The definition of a system, then, is conditional to the existence of suitable processes.

What still needs to be clarified is the difference between the independence from other systems of the system coming in or of the process itself. Suppose we have system $X$ and $Y$ whose state is identified by a single continuous variable. If there is no correlation, all state combinations are possible, so the joint distribution is uniform over the whole plain. Now suppose they are correlated so that that $x=-y$. The joint distribution is uniform on the diagonal. If we have a process that acts and depends only on $X$, the two cases are indistinguishable since the marginal of $x$ is uniformly distributed in both cases. The density of evolutions, however, is different, so there is a renormalization. But for the rest, the two cases are indistinguishable. So independence may not technically always map to factorization of the measure, but it may need to be indistinguishable from factorization.

\textbf{Independence leads to maximization of entropy}. Note that the presence of correlations between the system and the environment can only remove possible evolutions, not add them. Therefore independence is the case where the evolution count is maximal. In other word, requiring independence is equivalent to requiring maximization of entropy. Again note that this maximization of entropy is not due to some property of nature: it is due to the requirement of wanting to define a system, an objects for which we can mix and match processes. Of all processes nature presents us, we select those that decouple the world into separate systems, those that maximize entropy.

\textbf{Relaxation processes and equilibria as independence}. If we are to study systems, then, we must have at our disposal a process that takes an object and removes all possible correlations, rendering it independent. Moreover, this process must keep the system independent if it was already so. Therefore \textbf{independent descriptions must be equilibria} of such process. In general, the definition of the system requires the existence of such processes, which we call \textbf{relaxation} processes. States will be invariant, symmetries, of such processes.

\textbf{System definition as object-environment relationship}. Whether or not a particular set of attributes of an object can be seen as an independent system is not an a priori feature of the object, but rather a relationship between object and environment. This is not just because of independence, but also for the existence of the system itself. We can talk about a ball, with its position and velocity, on the surface of the earth at 20 Celsius. On the surface of the sun, we cannot talk about a ball. Therefore there must be a tight relationship between states, the processes defined on the system and the object-environment boundary. The process cannot alter the boundary, the state space must satisfy the constraints imposed by the boundary, the state space must be, as a whole, a symmetry of the processes.

\textbf{System composition and its algebra}. As we may want to describe systems, subsystems and their relationships, it useful to note that \textbf{the set of all possible systems is a Boolean algebra}. The ordering of the lattice is given by system inclusion. We say $X \in Y$ if $X$ is a subsystem of $Y$. We can convince ourselves that this is a partial order (i.e. $X \in X$, if $X \in Y$ and $Y \in X$ then $X=Y$ and if $X \in Y$ and $Y \in Z$ then $X \in Z$). The join $X \vee Y$ is therefore the smallest system that contains both and represents system composition. The meet $X \wedge Y$ is the biggest system contained by both and represent the common subsystem. The $1$ represents the system that contains everything. The $0$ represent the null system, the one that contains nothing. The complement $\neg X$ represents everything except for the system.

\textbf{Carving up the world}. Philosophers talk about a description of reality that is joint-carving meaning that reality already contains divisions, prior to our investigation. This is an epistemological version of it: nature gives us access to processes for which some systems are independent from the rest. It is the accessibility of these processes that allows us to define systems, without them we cannot divide nature into parts. That is, if an object is really made of parts but the process driving their evolution is such that the correlations can never be eliminated or controlled, the parts cannot be defined as independent systems.

\textbf{States and state spaces}. A \textbf{state is a snapshot for which the system is independent}. That is, it is a description of the system and only of the system. A \textbf{state space} $X$ is the set of all possible states of the system.

\textbf{States as junction points}. States will then act as the juncture point between processes. If we switch from process $A$ to process $B$ at time $t_0$, given the requirement of independence, the description of the system at that time must be a state. If that's the case, both processes can be characterized independently from each other.

\textbf{States as equilibria of faster/smaller scale processes}. As noted above, the subdomain must choose both a physical scale and a time scale. The requirement of independence will mean that the description within that scale is not relevant to the processes at hand. This cutoff is, in effect, part of the boundary of the system: we are still defining what descriptions, which degrees of freedom, are under consideration and which are not. Note that this is consistency of the idea of relaxation process, which must therefore not only remove the correlations from other objects but also from internal degrees of freedom.

\textbf{States as bundles of evolutions over short scale}. In this light, is it better to understand states not as points at an instant of time, but rather as bundles of evolutions over short scale that we ``connected'' throughout the process.

\textbf{State entropy}. When we join two processes, we need to reconcile the evolution count at the junction. This means that the evolution count for states must be well defined and unique if we want to mix and match antecedent and subsequent processes. Note that independence is precisely the condition needed for the count of evolutions to factorize and process entropy to becomes additive. In this case, we can define a \textbf{state entropy} $S(x)$ which is the process entropy in condition of system independence. This means that \textbf{all state spaces must be endowed with a structure that define the state entropy}.

\textbf{The entropic nature of geometric structures of state spaces}. The claim is that all geometric structures that state spaces possess in different theories are precisely structures that keep track of the state entropy. This is obvious in thermodynamics, where the equation of state in entropric form precisely map state entropy. In classical mechanics, phase space is characterized by a symplectic structure, whose volume is used in statistical mechanics to compute the entropy. Moreover, orthogonality in the symplectic structure characterize the independence of degrees of freedom. In quantum mechanics, the inner product is connected to probabilities that, as we saw, are simply ratios of counts of evolutions. Moreover, those same probabilities are used in statistical mechanics for the computation of the entropy. Also note that these structure do not provide further characterization of the state space beyond the ability to compute the evolution count and the entropy. They provide entropic and only entropic information.

\textbf{Zero state entropy and the null state}. As we saw before, entropy is a relative concept and therefore needs a unit to be defined. What state should act as a reference and be assigned zero entropy? There is a natural choice. Compositing with the null system $X \vee 0 = X$ returns the original system, therefore the entropy must not change. Since the entropy is linear under system composition, the null system must have zero entropy. We should note that, in principle, all systems should have a null state, which is the state where the system is not there. This case is automatically independent and is therefore a state. This state should be given zero entropy and therefore can be taken as the unit of our measure for the state space.

\textbf{Null state as the lowest entropy state}. It should also be intuitive that this state has to have the lowest entropy of all. That is, saying the system is not there will give the finest level of description of the system. No other description can be finer (except for the impossibility). This means that any other state can give a description of the system that is, at best, as fine as the null state. This means that the \textbf{the state entropy can never be negative}. This is a more general form of the third law of thermodynamics, in the same way that \ref{entropy_determinism} is a more general form of the second law.

\textbf{State entropy and time dependence}. We should stress that the state entropy fixes the relative process entropy between states at equal time in conditions of independence. In all other cases, or if these conditions changes throughout the process, the connection between state entropy and process entropy is more complicated: the same description for the same system will not correspond to the same count of evolutions. Consider a dynamical system with an attractor, such as a damped harmonic oscillator. As time proceeds, more and more evolutions will converge to the attractor. The same finite region around the attractor will correspond to more and more evolutions. In general, a process may introduce correlations between different variables which influences the value of the process entropy for a specific description.

\textbf{State space as template}. We should think of the state space as a template. We have a \textbf{state domain} $\tdomain_{X}$ which contain all statement fragments like \statement{the position of the ball is between 2 and 3 meters} together with a surjective map $\iota : \tdomain_{X} \times T \twoheadrightarrow \tdomain_t$ that adds \statement{at time t} for each possible time. The map $\iota$ must be a surjection that preserves the logical structure (i.e. logical operations, verifiability, narrowness, compatibility, ...) so that the basic logical relationships determined by the system itself are are valid at all times (i.e. \statement{at time t the position of the ball is between 2 and 3 meters} $\narrower$ \statement{at time t the position of the ball is between 0 and 300 meters} for every $t$).

Note, though, that $\iota( \cdot, t_0)$ in general is not an isomorphism (i.e. a bijection that preserves the logical structure) for two reasons. The first is that not all states will be available at every moment in every process. As we said before, if the process is dissipative the set of possible states will become smaller. In that case, $\iota$ will map all the inaccessible states to $\contradiction$, as they will be impossible. Therefore $\iota$ is not in general a bijection. The second reason is that the statement fragments in $\tdomain_{X}$ represent only descriptions about the system itself and nothing else. In a particular process, however, knowing the state of one system might tell us something about other system as well. The presence of correlations/coupling may mean that the precise knowledge of the degrees of freedom of the system may allow us to tell something about the environment or the internal structure (i.e. microstate). Mathematically, $\iota$ may map to statements that are narrower than the original fragment (e.g. we can have $\iota($\statement{position of A is 1 meter}$, t_0) \equiv$ \statement{at time $t_0$ the position of A is 1 meter and the position of B is -1 meter}).

\textbf{State spaces are constructed}. Though state spaces are template, conceptually they are not the starting point. The state space of the system (i.e. the set of all possible complete descriptions of the system and only about that system, independently of the process at hand) is something we need to construct. We find processes for which the system decouples, and we elevate those descriptions to independent objects. This construction comes with a set of assumptions about the system which are at the foundations of thermodynamics specifically and physics more in general, and they need to be made explicit. The snapshots (i.e. best descriptions of a system at a time), instead, do not require any additional assumptions and therefore are more general concepts.

\textbf{A rich mathematical structure} To sum up, the definition of a system, then, must define several things.
\begin{itemize}
	\item A boundary, that tells us where the system ends and the environment begins
	\item A characterization of the boundary, that defines in what cases the system remains well defined
	\item A set of relaxation processes, that tells us how the system can be decoupled from the rest
	\item A set of possible independent descriptions (i.e. states), that can be characterized about the system given the circumstances
	\item A set of possible evolutions for the system (i.e. the processes) under the given circumstances.
\end{itemize}
There is a lot more work to be done to fully characterize this mathematical structure.

%\section{Current attempt}

%\section{Examples}

%\section{Insightful failures}
